Comparative RNA-Seq: Signal to Noise Documentation

### 2022-03-06 manuscript.Rmd
large number of changes made. Investigating why render_toc is not found

---
title: "comparative_RNA-Seq_manuscript"
author: "Thomas_Nicholson"
date: "19/02/2022"
output: pdf_document
editor_options: 
  chunk_output_type: console
---

@@

@@

```{r upsetr_tree}
#ML phylogenetic tree built using 16s rRNA for the Gammaproteobacteria
#The full tree was also built for all bacterial genera, but is not displayed
tree <- read.tree(paste0(data_path, 'upsetr.tree'))

@@

```{r upsetr_plot}
#matrix of bool values indicating for each RNA if it was found in a genus
load(paste0(data_path, "upsetSubsetPC.Rda"))
load(paste0(data_path, "upsetSubsetPredicted.Rda"))

@@

```{r expressed_regions_frequency_plot}
#matrix of bool values indicating for each RNA if it was found in a genus
load(paste0(data_path, "upsetSubsetPC.Rda"))
load(paste0(data_path, "upsetSubsetPredicted.Rda"))

@@

```{r random_forest_frequency_plots}
load("~/bin/PhD/Chapter_4/chapter_4_files/predDat.Rda")
load("~/bin/PhD/Chapter_4/chapter_4_files/validatation2.Rda")

@@

```{r random_forest_statistics}
load("~/bin/PhD/Chapter_4/chapter_4_files/predDat.Rda")
load("~/bin/PhD/Chapter_4/chapter_4_files/validatation2.Rda")

@@

```{r statistics_plots}
load("~/bin/PhD/Chapter_4/chapter_4_files/predDat.Rda")
load("~/bin/PhD/Chapter_4/chapter_4_files/validatation2.Rda")

@@

```{r repeat_of_cumulative_distributions}
#is there a difference between low scoring (p < 0.17 and high scoring p > 0.5) predicted rufs?

>>>>

---
title: 'Comparative RNA-Seq: Signal to Noise Documentation'
author: "T.J Nicholson & P.P Gardner"
date: "March 2022"
output:
  html_document:
    toc: yes
    toc_depth: 4
    toc_float: yes
    number_sections: yes
  html_notebook:
    toc: yes
    toc_depth: 4
  pdf_document:
    toc: yes
    toc_depth: '4'
editor_options:
  chunk_output_type: console
---

@@

```{r toc, echo=FALSE, include=F} 
render_toc("manuscript.Rmd")
```

# Data Selection {#data_selection}

## Strains and experiments
How to get the summary table for the genera for the RNA-seq data:

-   Search the [SRA](https://www.ncbi.nlm.nih.gov/sra/) section of ncbi with 
    the genus name and ['Organism'] tag.

-   Select "Send results to Run selector"

-   Select the "Metadata" button of the "Total" row and "Download" Column.

-   The genome accession for each added genome was looked up on 
    [NCBI](https://www.ncbi.nlm.nih.gov/) using the strain name in the file or 
    by search for the given experiment in the 
    [SRA](https://www.ncbi.nlm.nih.gov/sra/) section.

-   Create a folder for the genus e.g. `mkdir Escherichia`

-   Create a folder using the GenBank assembly 
    accession e.g. `mkdir GCA_002504285.1.data/`

-   Create an `experiments_list.txt` file listing all of the SRA experiments 
    for the given genome and save the file in the `genus` folder.

-   Run `callPeaksforGenome.sh -g <GCA Accession> -s` from within 
    the `genus` folder.

-   List of representative genomes selected from RefSeq95
    -   2 genomes from each genera were selected (where possible)

## Predicting sRNAs

-   The `callPeaksforGenome.sh` carried out a number of steps
    -   `fasterq-dump <SRA Accession>` downloaded each of the RNA-Seq files
        listed in the `experiments_list.txt` file.
    -   `fetch_genomes_from_GCA.sh` downloaded the nucleotide FASTA file and
        the annotation file in GFF3 format.
    -   Known ncRNAs are identified using `cmscan` with models from Rfam
    -   `sra2plot.1.0.3.sh` took these files and produced a plot file (showing
        read depths at each nucleotide position) for each of the experiments.
    -   `run_rnaPeakCalling.R` took these plot files and identfied regions of
        RNA expression tht occured in intergenic regions
    -   `combine_gff_files.R` took the expressed regions from each of the 
        individual files (including the annotation file) and merged regions
        where there was an overlap.

-   The resulting regions are separated into `Known sRNA` and `Putative RUFs`
    based on whether an expressed region overlapped with a previously 
    identified ncRNA

-   The negative control group (RINCs) was generated 
    using `get_random_srna_sequences.py`
      -   For each putative RUF, a region of the same size was selected at 
          random from an intergenic region in the same genome

```{bash get_random_srna_sequences_loop, eval = F}
for file in *.txt; do 
  accession=`basename $file _new_calls.txt`; 
  get_random_srna_sequences.py -a $accession; 
  done
```

- The nucleotide sequences for each group were obtained from the nucleotide
  FASTA files using the coordinates of the regions

## Analysis

### Homology Search

-   `run_sRNA_nhmmer.sh -d <group_seqs.fna> -f <path/to/sequences/> -e fna`
    -   The `-e` flag is for the the expected extension of the input files.
    -   This loops through all the individual sequences in each data set and
        carries out a similarity search against all the sequences in the data
        set.
    -   The matched sequences are removed from the remaining search list to 
        prevent multiple identical alignments being produced.

-   `run_sRNA_nhmmer.sh -d <path/to/genomes/> -f <path/to/initial_alignment> -o` 
    `<path_to_output>/ -e stk -E 1e-3`
    -   The `-E` flag sets the e-value threshold for inclusion.
    -   This gets repeated two times to generate deep alignments for the 
        sequences from each data set.

-   A check was carried out for each of the groups (known sRNAs, predicted 
    RUFs and RINCs) that none of the alignments shared sequences.
      -   RINCs sharing sequences with either of the other data sets were 
          removed.
      -   Predicted RUFs sharing sequences with the known sRNAs were removed.

```{bash, eval=F}
group='predicted' #repeat for each group
cd ${group}/alignments/
> ../${group}_contig_pos.txt
for file in *.stk;
do 
  ID=`basename $file .stk`

  # extract the contig name and position for each sequence in each stockholm
  #alignment file
  grep "/" $file | grep -v "//" | cut -d ' ' -f1 | sed 's/\// /g' | 
  sed 's/-/ /g'| cut -d '|' -f2 | sort | uniq | sed -e "s/$/ $ID/" >> 
  ../${group}_contig_pos.txt
done

```{r eval=F}
library(GenomicRanges)
ncdat <- read.table("negative_control_contig_pos.txt", sep = " ", fill = T)

pcdat <- read.table("positive_control_contig_pos.txt", fill = T)

preddat <- read.table("predicted_contig_pos.txt", fill = T)

pcdat <- reformatContigPositionData(dat = pcdat)
ncdat <- reformatContigPositionData(dat = ncdat)
preddat <- reformatContigPositionData(dat = preddat)

nc.pc.Dat <- getOverlapIDs(queryDat =ncdat, targetDat = pcdat)
nc.pred.Dat <- getOverlapIDs(queryDat =ncdat, targetDat = preddat)
pc.pred.Dat <- getOverlapIDs(queryDat =preddat, targetDat = pcdat)

dat <- data.frame(ids = c(nc.pc.Dat$id1, nc.pred.Dat$id1))

write.table(dat, file = "not_negative_control_ids.txt", row.names = F, col.names = F, quote = F)

dat2 <- data.frame(ids =  pc.pred.Dat$id1)

write.table(dat2, file = "not_predicted_ids.txt", row.names = F, col.names = F, quote = F)

-   A similar check was carried out within each group to reduce redundancy
    -   Alignments sharing a sequence were merged into a single alignment.

```{bash, eval=F}

### Evolutionary Distance

-   A `cmscan` of all the genomes using the RF00177 Rfam model (bacterial 
      ribosomal SSU) was carried out.
-   [Phylip](https://evolution.genetics.washington.edu/phylip.html) was used to 
    build a maximum likelihood phyogenetic tree.
      -   The `RF00177.stk` alignment was reformatted to `.phylip` format
      -   `esl-reformat phylip RF00177.stk > RF00177.phylip`
      -   This was used as the input for `dnadist` with an output of 
          `RF00177.dnadist`.

@@

```{r upsetr_tree, eval=F}
#ML phylogenetic tree built using 16s rRNA for the Gammaproteobacteria
#The full tree was also built for all bacterial genera, but is not displayed
tree <- read.tree(paste0(data_path, 'upsetr.tree'))

@@

```{r upsetr_plot, eval=F}
#matrix of bool values indicating for each RNA if it was found in a genus
load(paste0(data_path, "upsetSubsetPC.Rda"))
load(paste0(data_path, "upsetSubsetPredicted.Rda"))

@@

```{r expressed_regions_frequency_plot, eval=F}
#matrix of bool values indicating for each RNA if it was found in a genus
load(paste0(data_path, "upsetSubsetPC.Rda"))
load(paste0(data_path, "upsetSubsetPredicted.Rda"))

@@

  -   The output of the RF classifier for the test data and the predicted RUFs

```{r random_forest_frequency_plots, }
load("~/bin/PhD/Chapter_4/chapter_4_files/predDat.Rda")
load("~/bin/PhD/Chapter_4/chapter_4_files/validatation2.Rda")
#select the desired columns from the predicted data and validatation data

probDat <- predDat %>% 
  select(probability, ID, group, srna.counts.2) %>% 
  bind_rows(validation2 %>% select(probability, ID, group, srna.counts.2))

@@

  -   The predicted RUFs were split into <strong>transcriptional noise</strong>, 
      <strong>possible RNA</strong>, and <strong>putative RNA</strong>.
  -   The categories were then compared to see if there was a difference in the
      level of transcription between the groups.

```{r transcription_levels_by_result}

```{r random_forest_statistics, eval=F}
#TODO this is not used
load("~/bin/PhD/Chapter_4/chapter_4_files/predDat.Rda")
load("~/bin/PhD/Chapter_4/chapter_4_files/validatation2.Rda")

@@

```{r statistics_plots, eval=F}
#TODO this is not used
load("~/bin/PhD/Chapter_4/chapter_4_files/predDat.Rda")
load("~/bin/PhD/Chapter_4/chapter_4_files/validatation2.Rda")

@@

```{r repeat_of_cumulative_distributions, eval=F}
#TODO this is not used
#is there a difference between low scoring (p < 0.17 and high scoring p > 0.5) predicted rufs?
